Maximum likelihood estimation of cardinalities\\ from HyperLogLog sketches

### 2016-05-01 paper/paper.tex
added some paragraphs to the result section

\autoref{fig:estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \algoref{alg:estimate_ml} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$. As the median shows, the error is essentially unbiased over the entire cardinality range except for very small values. The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is an order of magnitude larger than the operating range upper bound of the \emph{raw} estimate (\autoref{fig:raw_estimate}) or the original method (\autoref{fig:original_estimate}).

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_8_24}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:estimation_error_8_24}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_14}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:estimation_error_12_14}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:estimation_error_16_16}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_20_12}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 20$ and $\symRegRange=12$.}
\label{fig:estimation_error_20_12}
\end{figure}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:estimation_error_12_52}
\end{figure}

\autoref{fig:estimation_error_8_24}
\autoref{fig:estimation_error_12_14}
\autoref{fig:estimation_error_12_20}
\autoref{fig:estimation_error_12_52}
\autoref{fig:estimation_error_16_16}
\autoref{fig:estimation_error_20_12}

>>>>

\autoref{fig:estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \algoref{alg:estimate_ml} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$. As the median shows, the error is essentially unbiased over the entire cardinality range except for very small values. The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the \emph{raw} estimate (\autoref{fig:raw_estimate}) or the original method (\autoref{fig:original_estimate}).

@@

The new estimation algorithm also works well for other HyperLogLog configurations. First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$). The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=20$ are shown in \autoref{fig:estimation_error_8_24}, \autoref{fig:estimation_error_16_16}, and \autoref{fig:estimation_error_20_12}, respectively. As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision. Again, the new algorithm gives inherently unbiased estimates.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_8_24}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:estimation_error_8_24}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:estimation_error_16_16}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_20_12}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 20$ and $\symRegRange=12$.}
\label{fig:estimation_error_20_12}
\end{figure}

As proposed in \cite{Heule2013}, the operating range can be extended by
replacing the 32-bit hash function by a 64-bit hash function. \autoref{fig:estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$. In this case, in order to use all the 64 bits of the hash value, $\symRegRange$ must be chosen to be equal to $64 - \symPrecision = 52$. As a consequence, in order to represent the maximum possible register value $\symRegRange+1 = 53$, 6 bits are needed for each register. The doubled hash value size shifts the maximum trackable cardinality value towards $2^{64}$. As \autoref{fig:estimation_error_12_52} shows, when compared to the 32-bit hash value case given in \autoref{fig:estimation_error_12_20}, the estimation error remains constant over the entire simulated cardinality range up to 50 billions.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:estimation_error_12_52}
\end{figure}

\autoref{fig:estimation_error_12_14}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_14}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:estimation_error_12_14}
\end{figure}
